GeoSPARQL 1.1: an almost decadal update to the most important geospatial LOD standard

### 2021-03-12 manuscript.tex
reduced to 7 pages

\author{
    Nicholas J. Car\inst{1}\orcidID{0000-0002-8742-7730} \and \\
    Timo Homburg\inst{2}\orcidID{0000-0002-9499-5840} \and \\
    Simon J.D Cox\inst{3}\orcidID{0000-0002-3884-3420}
}

@@

\institute{
    {
    SURROUND Australia Pty Ltd., Australia \&\\
    Australian National University, Australia\\
    \email{nicholas.car@surroundaustralia.com}\\
    \url{https://surroundaustralia.com}
    }
    \and
    {
        Mainz University Of Applied Sciences, Germany\\
        \email{timo.homburg@hs-mainz.de}
    }
    \and 
    {
        Commonwealth Scientific \& Industrial Research Organisation, Australia\\
        \email{simon.cox@csiro.au}
    }
}

@@

\begin{abstract}
The Open Geospatial Consortium published the GeoSPARQL 1.0 standard in 2012 containing multiple 
parts that define ``SPARQL extension functions'', ``RIF rules'', ``an RDF/OWL ontology for 
information based on the General Feature Model'' and supporting vocabularies, all for Semantic 
Web spatial data.

In the 8+ years since its publication, GeoSPARQL has become the most important spatial Semantic 
Web standard, as judged by references to it in other Semantic Web standards and its wide use in 
Semantic Web data.

An update to the standard was proposed in 2019 to deliver GeoSPARQL 1.1 in 2021 with a charter to: 
handle outstanding change requests and source new ones from the GeoSPARQL user community as well 
to ``better present'' the standard, that is to better link all the standard’s parts and better 
document \& exemplify elements. Expected updates included possible alignments to other ontologies, 
possible handling of new spatial referencing systems, new geometry representations, and new artifact 
presentation.

@@

\section{Introduction}\label{sec:introduction}
The GeoSPARQL standard, first issues in 2021 by the Open Geospatial Consortium (OGC)\footnote{\url{https://www.ogc.org}} 
is one of, if not the most\footnote{It's hard to calculate use but references to GeoSPARQL other well-known standards, 
such as DCAT2 (\url{https://www.w3.org/TR/vocab-dcat/}) suggests this}, used \textit{Sematic Web} ontologies for 
representing spatial data.

The original GeoSPARQL release, which we refer to as GeoSPARQL 1.0, contained a \textit{specification} document,
a main ``GeoSPARQL'' ontology in an RDF file and a ``Simple Features Vocabulary'' ontology also in an RDF file. The 
``GeoSPARQL'' ontology content, as well as lists of geospatial functions that could be performed on RDF data via 
SPARQL\footnote{\url{https://www.w3.org/TR/sparql11-query/}} queries was defined in the specification document, as
were entailment rules and requirements \& abstract tests for testing ontology data and function implementations. 
Within the last few years, the function lists from the specification were extracted into a SKOS\footnote{\url{https://www.w3.org/TR/skos-reference/}}
vocabulary.

\section{Motivation to update GeoSPARQL}\label{sec:motivation}
The Open Geospatial Consortium (OGC) \& World Wide Web Consortium's (W3C) \textit{Spatial Data On The Web Working Group}
(SDWWG) established \textit{Spatial Data On The Web Best Practices}\cite{van_den_brink_best_2018} which noted 
shortcomings with then current spatial data standards such as ``A best practice for returning geometries in a 
specific requested CRS has not yet emerged''. The group also informally captured specific suggested updates for 
GeoSPARQL\footnote{\url{https://www.w3.org/2015/spatial/wiki/Further_development_of_GeoSPARQL}}, however no updates 
to GeoSPARQL were ultimately made by that Working Group.

Recently, 2019, the OGC reconstituted a \textit{GeoSPARQL Standards Working Group} (SWG) to update GeoSPARQL. The general 
motivation for work within the area of GeoSPARQL, that of \textit{Semantic Web} spatial data, and a series of
fault fixes and proposed extensions to GeoSPARQL 1.0 are captured in an OGC White Paper \cite{geosparqlwhitepaper}. Some,
but not all of the SDWWG's ideas have been taken up by the SDW, for example the \textit{Best Practices}~\cite{van_den_brink_best_2018}
aspiration that ``A possible way forward is an update for the GeoSPARQL spatial ontology. This would provide an 
agreed spatial ontology, i.e., a bridge or common ground between geographical and non-geographical spatial data...''.
This point has been considered by the SWG and included in future releases' scope.

Another \textit{Best Practices} issue raised is that ``it makes sense to publish different geometric representations 
of a spatial object that can be used for different purposes'', again, this is being considered by the SWG with initial
thoughts cenetering on defining \textit{roles} of geometries with respect to features.

@@

The first two of these new serializations have been much anticipated and were requested by the SDWWG and many users of 
GeoSPARQL, due to those formats' popularity.

The third new serialization format is more forward-looking in that it is not driven by user demand but by predicted demand.
DGGS does not have a single, concrete format standard as the others do, nor is it ever likely to - different DGGSes will 
likely implement very different data formats - so GeoSAPRQL 1.1 makes generalized provisions for DGGS serializations but 
presents no detailed requirements for them, only stating that the specific DGGS must be identified.

@@

New functions to retrieve the minimum and maximum values of geometries' coordinates are also added: \href{http://www.opengis.net/def/function/geosparql/minX}{\emph{geof:minX}} \& \href{http://www.opengis.net/def/function/geosparql/maxX}{\emph{geof:maxX}},
\href{http://www.opengis.net/def/function/geosparql/minY}{\emph{geof:minY}} \& \href{http://www.opengis.net/def/function/geosparql/maxY}{\emph{geof:maxY}} and \href{http://www.opengis.net/def/function/geosparql/minZ}{\emph{geof:minZ}} \& \href{http://www.opengis.net/def/function/geosparql/maxZ}{\emph{geof:maxZ}}.

\subsection{Ontology extensions}\label{sec:ontexts}
Figure~\ref{fig:geosparql11ontology} overviews the ontology, as 
of release 1.1.

GeoSPARQL 1.1 extends the GeoSPARQL ontology by adding a new class, \href{http://www.opengis.net/ont/geosparql#SpatialMeasure}{\texttt{geo:SpatialMeasure}}. This class represents a spatial measurement 
such as a volume, length, or area associated with a measurement amount and a unit of measure. It is the range of three newly-defined properties:
\href{http://www.opengis.net/ont/geosparql#hasArea}{\texttt{geo:hasArea}}, \href{http://www.opengis.net/ont/geosparql#hasLength}{\texttt{geo:hasLength}} and \href{http://www.opengis.net/ont/geosparql#hasVolume}{\texttt{geo:hasVolume}} which make these attributes of a geometry better accessible using 
SPARQL.

These additions address a number of change requests from the SDWWG and the SWG but, given the introduction of a dedicated \textit{measure} class, 
the do more than just provide for a few scalar spatial measures and open up GeoSPARQL to general patterns of measurement present in ontologies 
such as the W3C's \textit{SOSA}~\cite{haller_semantic_2017}.

The 1.1 release also adds the property \href{http://www.opengis.net/def/function/geosparql/inSRS}{\texttt{geo:inSRS}}, which allows declarations of a geometry's SRS, independent of serializations. It paves the way 
for future definition of SRSes in RDF, as anticipated for GeoSPARQL 2.0.

@@

\section{Conclusions}\label{sec:conclusions}
An staged schedule of updates to this very important \textit{Semantic Web} spatial data standard has been initiated with simple, objvious and strictly backwards-compatable changes now in progress for GeoSPARQL 1.1. 1.2 will include more detailed work later in 2021 and 2.0, as yet un-specified, might be a somewhat different stanandard, to be determined by more requirements gathering and feedback on the 1.1 and 1.2 releases.

>>>>

\author{
    Nicholas J. Car\inst{1}\orcidID{0000-0002-8742-7730} \and \\
    Timo Homburg\inst{2}\orcidID{0000-0002-9499-5840}\\
%    Simon J.D Cox\inst{3}\orcidID{0000-0002-3884-3420}
}

@@

\institute{
    {
    %SURROUND Australia Pty Ltd., Australia \&\\
    Australian National University, Australia\\
    \email{nicholas.car@surroundaustralia.com}\\
    %\url{https://surroundaustralia.com}
    }
    \and
    {
        Mainz University Of Applied Sciences, Germany\\
        \email{timo.homburg@hs-mainz.de}
    }
    % \and 
    % {
    %     Commonwealth Scientific \& Industrial Research Organisation, Australia\\
    %     \email{simon.cox@csiro.au}
    % }
}

@@

\begin{abstract}
In 2012 the Open Geospatial Consortium published GeoSPARQL 
defining ``SPARQL extension functions'', ``RIF rules'', ``and RDF/OWL ontology for 
[spatial] information'' and supporting vocabularies.\\

In the 8+ years since its publication, GeoSPARQL has become the most important spatial Semantic 
Web standard, as judged by references to it in other Semantic Web standards and its wide use in 
Semantic Web data.\\

An update to the standard was proposed in 2019 to deliver GeoSPARQL 1.1 in 2021 with a charter to: 
handle outstanding change requests and source new ones from the user community and 
to ``better present'' the standard, that is to better link all the standard’s parts and better 
document \& exemplify elements. Expected updates included alignments to other ontologies, 
handling of new spatial referencing systems, new geometry representations, and new artifact 
presentation.\\

@@

\section{Introduction}\label{sec:introduction}
The GeoSPARQL standard, issued in 2012 by the Open Geospatial Consortium (OGC)\footnote{\url{https://www.ogc.org}} 
is one of, if not the most\footnote{It's hard to calculate use but references to GeoSPARQL other well-known standards, 
such as DCAT2 (\url{https://www.w3.org/TR/vocab-dcat/}) suggests this}, used \textit{Sematic Web} standards for 
spatial data.

The original release - GeoSPARQL 1.0 - contained a \textit{specification} document,
a main ``GeoSPARQL'' ontology in an RDF file and a ``Simple Features Vocabulary'' ontology also in an RDF file. The 
``GeoSPARQL'' ontology content, as well as lists of geospatial functions that could be performed on RDF data via 
SPARQL\footnote{\url{https://www.w3.org/TR/sparql11-query/}} queries were defined in the specification document, as
were entailment rules and requirements \& abstract tests for testing ontology data and function implementations. 
Function lists from the specification have been extracted into SKOS\footnote{\url{https://www.w3.org/TR/skos-reference/}}
vocabularies.

\section{Motivation to update GeoSPARQL}\label{sec:motivation}
The OGC \& World Wide Web Consortium's (W3C) \textit{Spatial Data On The Web Working Group}
(SDWWG) established \textit{Spatial Data On The Web Best Practices}\cite{van_den_brink_best_2018} which noted 
shortcomings with then current spatial data standards: ``A best practice for returning geometries in a 
specific requested CRS has not yet emerged''. The group also informally captured specific suggested updates for 
GeoSPARQL\footnote{\url{https://www.w3.org/2015/spatial/wiki/Further_development_of_GeoSPARQL}}, however no updates 
to GeoSPARQL were then made.

Recently, 2019, the OGC reconstituted a \textit{GeoSPARQL Standards Working Group} (SWG) to update GeoSPARQL. The general 
motivation for work within the area of GeoSPARQL, that of \textit{Semantic Web} spatial data, and a series of
fault fixes and proposed extensions to GeoSPARQL 1.0 are captured in an OGC White Paper \cite{geosparqlwhitepaper}. Some,
but not all, of the SDWWG's ideas have been taken up by the SDW, for example the \textit{Best Practices}~\cite{van_den_brink_best_2018}
aspiration that ``A possible way forward is an update for the GeoSPARQL spatial ontology. This would provide an 
agreed spatial ontology, i.e., a bridge or common ground between geographical and non-geographical spatial data...''.
This point has been considered by the SWG and included in future releases' scope.

Another \textit{Best Practices} issue raised is that ``it makes sense to publish different geometric representations 
of a spatial object that can be used for different purposes''. This is being considered by the SWG with initial
thoughts centering on defining \textit{roles} of geometries with respect to features.

@@

GeoJSON \& KML have been much anticipated and were requested by the SDWWG and many users of 
GeoSPARQL, due to those formats' popularity.

The DGGS format is more forward-looking in that it is not driven by user demand but by predicted demand.
DGGS does not have a single, concrete format standard as the others do, nor is it ever likely to - different DGGSes will 
likely implement very different data formats - so GeoSAPRQL 1.1 makes generalized provisions for DGGS serializations but 
presents no detailed requirements for them, only stating that the specific DGGS must be identified.

@@

Functions to retrieve min/max values of geometries' coordinates are added: \href{http://www.opengis.net/def/function/geosparql/minX}{\emph{geof:minX}} \& \href{http://www.opengis.net/def/function/geosparql/maxX}{\emph{geof:maxX}},
\href{http://www.opengis.net/def/function/geosparql/minY}{\emph{geof:minY}} \& \href{http://www.opengis.net/def/function/geosparql/maxY}{\emph{geof:maxY}} and \href{http://www.opengis.net/def/function/geosparql/minZ}{\emph{geof:minZ}} \& \href{http://www.opengis.net/def/function/geosparql/maxZ}{\emph{geof:maxZ}}.

\subsection{Ontology extensions}\label{sec:ontexts}
GeoSPARQL 1.1 - see Figure~\ref{fig:geosparql11ontology} for an overview - extends the GeoSPARQL ontology by adding a new class, \href{http://www.opengis.net/ont/geosparql#SpatialMeasure}{\texttt{geo:SpatialMeasure}}. This class represents a spatial measurement 
such as a volume, length, or area associated with a measurement amount and a unit of measure. It is the range of three newly-defined properties:
\href{http://www.opengis.net/ont/geosparql#hasArea}{\texttt{geo:hasArea}}, \href{http://www.opengis.net/ont/geosparql#hasLength}{\texttt{geo:hasLength}} and \href{http://www.opengis.net/ont/geosparql#hasVolume}{\texttt{geo:hasVolume}} which make these attributes of a geometry better accessible using 
SPARQL.

These additions address a several requests from the SDWWG \& SWG but, given the introduction of a dedicated \textit{measure} class, 
do more too: they open up GeoSPARQL to general patterns of measurement present in ontologies 
such as the W3C's \textit{SOSA}~\cite{haller_semantic_2017}.

The 1.1 release also adds the property \href{http://www.opengis.net/def/function/geosparql/inSRS}{\texttt{geo:inSRS}}, which allows declarations of a geometry's SRS, independent of serializations which paves the way 
for future definition of SRSes in RDF, as anticipated for GeoSPARQL 2.0.

@@

\section{Conclusions}\label{sec:conclusions}
A staged schedule of updates to this important \textit{Semantic Web} spatial standard has been initiated with simple and strictly backwards-compatable changes now in GeoSPARQL 1.1. 1.2 will include more detailed work later in 2021 and 2.0, as yet un-specified, might be somewhat different: to be determined by more requirements gathering and feedback on the 1.1 and 1.2 releases.
